---
title: "Population and Lyme Disease Prevalence in Maine"
output: 
  html_document:
     toc: TRUE
     number_sections: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r libraries, echo = FALSE}

# install packages

library(tidyverse) 
library(leaflet) 
library(sf) 
library(RColorBrewer) 
library(htmltools) 
library(leafsync) 
library(kableExtra) 
library(tidycensus)
library(rvest)
library(DT)
library(fedmatch)

```

# Introduction

In order to determine the regions of Maine with the greatest risk associated with Lyme disease, it is important to understand where Lyme disease cases are most often reported as well as how the population of Maine is distributed across the state. While it is nearly impossible to gather tick population data, comparing human population numbers to Lyme disease prevalence data may provide valuable insight into where ticks tend to congregate as well as the risk of contracting Lyme disease in different parts of the state.

# Data

## Tickborne Diseases

Lyme disease prevalence data from 2019-2022 for the state of Maine is publicly available on the University of Maine website. Data for several other tickborne diseases is also available on the website.


### UMaine 2019 Season Prevalence Data By Town {.unlisted .unnumbered}

```{r data-import-umaine-prevalence, echo = FALSE}

# read csv file
tick_towns <- read_csv("/cloud/project/data/ticks/umaine_tickborne_prevalence_town.csv")

```

```{r umaine-tick-prevalence-data-table}

# construct interactive data table displaying UMaine tick testing prevalence data
DT::datatable(tick_towns,
              options = list(
                paging = TRUE,
                pageLength = 10),
              height = 500)

```

### Maine Tracking Network Rate of Tickborne Disease by Town 2016-2020 {.unlisted .unnumbered}

```{r data-import-mtn-rate, echo = FALSE}

# read csv file
mtn_tick_town_rate <- read_csv("/cloud/project/data/ticks/maine_tracking_network_rate.csv")

# reformat rate
mtn_tick_town_rate[mtn_tick_town_rate == "*"] = NA
mtn_tick_town_rate[mtn_tick_town_rate == "NR"] = NA

```

```{r mtn-tick-rate-data-table}

# construct interactive data table displaying tick rate data
DT::datatable(mtn_tick_town_rate,
              options = list(
                paging = TRUE,
                pageLength = 10),
              height = 500)

```

### Number of Tickborne Disease Cases by Town 2016-2020 {.unlisted .unnumbered}

```{r data-import-mtn-number, echo = FALSE}

# read csv file
mtn_tick_town_number <- read_csv("/cloud/project/data/ticks/maine_tracking_network_incidence.csv")

# reformat count variables
mtn_tick_town_number <- mtn_tick_town_number %>%
  mutate(Lyme_Label = Lyme) %>%
  mutate(Lyme_Label = str_replace(Lyme_Label, "NR", "Not Releasable")) %>%
  mutate(Lyme = str_replace(Lyme, "<6", "6")) %>%
  mutate(Lyme = str_replace(Lyme, "6-10", "8")) %>%
  mutate(Lyme = str_replace(Lyme, "11-15", "13")) %>%
  mutate(Babesiosis_Label = Babesiosis) %>%
  mutate(Babesiosis_Label = str_replace(Babesiosis_Label, "NR", "Not Releasable")) %>%
  mutate(Babesiosis = str_replace(Babesiosis, "<6", "6")) %>%
  mutate(Babesiosis = str_replace(Babesiosis, "6-10", "8")) %>%
  mutate(Babesiosis = str_replace(Babesiosis, "11-15", "13")) %>%
  mutate(Anaplasmosis_Label = Anaplasmosis) %>%
  mutate(Anaplasmosis_Label = str_replace(Anaplasmosis_Label, "NR", "Not Releasable")) %>%
  mutate(Anaplasmosis = str_replace(Anaplasmosis, "<6", "6")) %>%
  mutate(Anaplasmosis = str_replace(Anaplasmosis, "6-10", "8")) %>%
  mutate(Anaplasmosis = str_replace(Anaplasmosis, "11-15", "13"))
  

# create numeric count variables
mtn_tick_town_number$Lyme <- as.numeric(mtn_tick_town_number$Lyme)
mtn_tick_town_number$Babesiosis <- as.numeric(mtn_tick_town_number$Babesiosis)
mtn_tick_town_number$Anaplasmosis <- as.numeric(mtn_tick_town_number$Anaplasmosis)

mtn_tick_town_number[mtn_tick_town_number == "NR"] = NA

# reformat location variable
mtn_tick_town_number <- mtn_tick_town_number %>%
  mutate(Location = str_replace(Location, "Plt", "Plantation")) %>%
  mutate(Location = str_replace(Location, "Bancroft Twp", "Bancroft")) %>%
  mutate(Location = str_replace(Location, "Aroostook", "Aroostook Twp")) %>%
  mutate(Location = str_replace(Location, "Somerset", "Somerset Twp")) %>%
  mutate(Location = str_replace(Location, "East Hancock", "East Hancock Twp")) %>%
  mutate(Location = str_replace(Location, "Dennistown", "Dennis")) %>%
  mutate(Location = str_replace(Location, "East Central Washington", "East Central Washington Twp")) %>%
  mutate(Location = str_replace(Location, "East Central Franklin", "East Central Franklin Twp")) %>%
  mutate(Location = str_replace(Location, "West Central Franklin", "West Central Franklin Twp")) %>%
  mutate(Location = str_replace(Location, "North Franklin", "North Franklin Twp")) %>%
  mutate(Location = str_replace(Location, "South Franklin", "South Franklin Twp")) %>%
  mutate(Location = str_replace(Location, "West Franklin", "West Franklin Twp")) %>%
  mutate(Location = str_replace(Location, "East Central Penobscot", "East Central Penobscot Twp")) %>%
  mutate(Location = str_replace(Location, "Louds Island", "Louds Island Twp")) %>%
  mutate(Location = str_replace(Location, "Marshall Island", "Marshall Island Twp")) %>%
  mutate(Location = str_replace(Location, "Monhegan Island Plantation", "Monhegan Plantation")) %>%
  mutate(Location = str_replace(Location, "Islands", "Islands Twp")) %>%
  mutate(Location = str_replace(Location, "North Oxford", "North Oxford Twp")) %>%
  mutate(Location = str_replace(Location, "South Oxford", "South Oxford Twp")) %>%
  mutate(Location = str_replace(Location, "North Washington", "North Washington Twp")) %>%
  mutate(Location = str_replace(Location, "North Penobscot", "North Penobscot Twp")) %>%
  mutate(Location = str_replace(Location, "Piscataquis", "Piscataquis Twp")) %>%
  mutate(Location = str_replace(Location, "Prentiss Twp T7 R3 NBPP", "Prentiss Twp")) %>%
  mutate(Location = str_replace(Location, "Seboomook Lake", "Seboomook Lake Twp")) %>%
  mutate(Location = str_replace(Location, "Square Lake", "Square Lake Twp")) %>%
  mutate(Location = str_replace(Location, "Saint", "St.")) 
  
```

```{r mtn-tick-number-data-table}

# construct interactive data table displaying tick incidence data
DT::datatable(mtn_tick_town_number,
              options = list(
                paging = TRUE,
                pageLength = 10),
              height = 500)

```

## Population 

Population data for the state of Maine is publicly available via the U.S. Census Bureau records. This data exists in several different spatial capacities, so determining the most useful spatial unit is very important when working with the data. The data itself can be imported using a unique API key in RStudio.

### Importing population data from the U.S. Census Bureau {.unlisted .unnumbered}

```{r census-data-import, echo = TRUE, warning = FALSE, message = FALSE, results = 'hide'}

# apply unique census api key
census_api_key("5177724b01a7fe4714097e711cb95230c37cfce7", install = TRUE, overwrite = TRUE)

# import census data 
## spatial unit --> place
## variable of interest --> population
census_place_df <- get_acs(geography = "place", variables = c("B01003_001E"), state = "ME", geometry = TRUE)

# import census data 
## spatial unit --> county subdivision
## variable of interest --> population
census_county_sub_df <- get_acs(geography = "county subdivision", variables = c("B01003_001E"), state = "ME", geometry = TRUE)

```

<br>

```{r reformat-location-variables, echo = FALSE}

# reformat location variables >> census place df
census_place_df <- census_place_df %>%
  separate(NAME, into = c("Place", "State"), sep = ",") %>%
  mutate(Place = str_remove(Place, " CDP| city"))

# reformat location variables >> census county subdivision df
census_county_sub_df <- census_county_sub_df %>%
  separate(NAME, into = c("Place", "County", "State"), sep = ",") %>%
  mutate(Place = str_remove(Place, "town| city")) %>%
  mutate(Place = str_replace(Place, "UT", "Twp")) %>%
  mutate(Place = str_replace(Place, "plantation", "Plantation")) %>%
  mutate(Place = str_replace(Place, "George town", "Georgetown")) %>%
  mutate(Place = str_replace(Place, "Hibberts gore", "Hibberts Gore")) %>%
  mutate(Place = str_trim(Place, side = "both")) %>%
  mutate(Place = str_replace(Place, "Passamaquoddy Pleasant Point Reservation", "Pleasant Point")) %>%
  mutate(Place = str_replace(Place, "Passamaquoddy Indian Township Reservation", "Indian Twp Res"))

# reformat location variables >> tick towns df
tick_towns <- tick_towns %>%
  mutate(town = str_replace(town, "plt", "Plantation")) %>%
  mutate(town = str_replace(town, "Plt", "Plantation")) %>%
  mutate(town = str_replace(town, "Monhegan Island", "Monhegan")) %>%
  mutate(town = str_replace(town, "Muscle Ridge", "Muscle Ridge Islands")) %>%
  mutate(town = str_replace(town, "Cary Twp", "Cary Plantation")) %>%
  mutate(town = str_replace(town, "Codyville Twp", "Codyville Plantation")) %>%
  mutate(town = str_replace(town, "Seboomook Twp", "Seboomook Lake Twp")) %>%
  mutate(town = str_replace(town, "Prentiss Twp T4 R4 NBKP", "Prentiss Twp")) %>%
  mutate(town = str_replace(town, "Dennistown", "Dennis")) %>%
  mutate(town = str_replace(town, "Saint", "St.")) %>%
  mutate(town = str_replace(town, "Atkinson Twp", "Atkinson")) %>%
  mutate(town = str_replace(town, "Bancroft Twp", "Bancroft")) 
```

### Census data using "place" spatial unit {.unlisted .unnumbered}

```{r census-data-table-place}

# construct interactive data table displaying census data (place)
DT::datatable(census_place_df,
              options = list(
                paging = TRUE,
                pageLength = 10),
              height = 500)

```

<br>

### Census data using "county subdivision" spatial unit {.unlisted .unnumbered}

```{r census-data-table-county-subdivision}

# construct interactive data table displaying census data (county subdivision)
DT::datatable(census_county_sub_df,
              options = list(
                paging = TRUE,
                pageLength = 10),
              height = 500)

```

### Spatial unit differences {.unlisted .unnumbered}

Differences in the size and coverage of data points from the Census data can be visualized using static spatial plots. Although the "place" data does a good job displaying the major population centers of Maine, the "county subdivision" data appears to have more complete coverage of the state.

```{r visualize-population-data}

# visualize polygons using ggplot
plot1 <- ggplot(census_place_df) +
  geom_sf(aes(fill = estimate)) +
  labs(fill = "Population",
       title = "Maine population",
       subtitle = "By census place")

# visualize polygons using ggplot
plot2 <- ggplot(census_county_sub_df) +
  geom_sf(aes(fill = estimate)) +
  labs(fill = "Population",
       title = "Maine population",
       subtitle = "By census county subdivision") 

gridExtra::grid.arrange(plot1, plot2, ncol=2)

```

```{r organize-data, echo = FALSE}

# set CRS
census_place_df <- st_transform(census_place_df, "+init=epsg:4326")
census_county_sub_df <- st_transform(census_county_sub_df, "+init=epsg:4326")

# join tick data with census data >> census place df with tick towns df
tick_census_place_df <- tick_towns %>%
  left_join(census_place_df, by = c("town" = "Place"))

# join tick data with census data >> census county subdivision df with tick towns df
tick_census_county_sub_df <- census_county_sub_df %>%
  left_join(tick_towns, by = c("Place" = "town"))

# view non-matches >> census county subdivision df with tick towns df
missing <- anti_join(census_county_sub_df, tick_towns, by = c("Place" = "town")) %>%
                    select(Place, County)

```

### Population and Tick Prevalence Leaflet {.unlisted .unnumbered}

The population data and the tick prevalence data can be explored relative to one another on the interactive leaflet plot shown below. It is interesting to see that the majority of the tick cases in the data are located on the coast, and not necessarily in the largest population centers in the state.

```{r tick-and-population-leaflet}

# create labels
pop_labels <- sprintf(
  "<strong>%s (2019)</strong> <br/>Population: %g",
   tick_census_county_sub_df$Place, tick_census_county_sub_df$estimate
) %>% lapply(htmltools::HTML)

tick_labels <- sprintf(
  "<strong>%s (2019)</strong> <br/>Positive Lyme Tests: %g, <br/> Positivity Rate: %s",
    tick_census_county_sub_df$Place, 
    tick_census_county_sub_df$borrelia_positive, 
    tick_census_county_sub_df$borrelia_percent
) %>% lapply(htmltools::HTML)

# create layer names
Groupnames <- c("Population", "Lyme Disease Prevalence")

# define bins
bins1 <- seq(0, 80000, by = 10000)
bins2 <- seq(0, 30, by = 3)

# select color palettes
pal1 <- colorBin("OrRd", domain = tick_census_county_sub_df$estimate, bins = bins1)
pal2 <- colorBin("PuRd", domain = tick_census_county_sub_df$borrelia_positive, bins = bins2)

# generate leaflet
leaflet(tick_census_county_sub_df) %>%
  addProviderTiles(providers$OpenStreetMap) %>%
  addPolygons(fillColor = ~pal1(tick_census_county_sub_df$estimate),
              color = "black",
              weight = 1,
              fillOpacity = 0.9,
              label = ~pop_labels,
              group = "Population") %>%
  addPolygons(fillColor = ~pal2(tick_census_county_sub_df$borrelia_positive),
              color = "black",
              weight = 1,
              fillOpacity = 0.9,
              label = ~tick_labels,
              group = "Lyme Disease Prevalence") %>%
  addLegend(values = ~tick_census_county_sub_df$estimate, 
            group = "Population", 
            position = "bottomleft", 
            title = "Population",
            opacity = 1,
            pal = pal1) %>%
  addLegend(values = ~tick_census_county_sub_df$babesia_positive, 
            group = "Lyme Disease Prevalence", 
            position = "bottomleft", 
            title = "Lyme Disease Prevalence",
            opacity = 1,
            pal = pal2) %>%
  setView(lng = -69.5, 
          lat = 45.3, 
          zoom = 6.5) %>%
  addLayersControl(
     baseGroups = c("Population", "Lyme Disease Prevalence"),
     options = layersControlOptions(collapsed = FALSE)) %>%
  htmlwidgets::onRender("
    function(el, x) {
      var updateLegend = function () {
          var selectedGroup = document.querySelectorAll('input:checked')[0].nextSibling.innerText.substr(1);
          document.querySelectorAll('.legend').forEach(a => a.hidden = true);
          document.querySelectorAll('.legend').forEach(l => {
            if (l.children[0].children[0].innerText == selectedGroup) l.hidden = false;
          });
      };
      updateLegend();
      this.on('baselayerchange', e => updateLegend());
    }")

```

```{r mtn-tick-leaflet}

# join tick data with census data >> census county subdivision df with mtn number df
mtn_number_census_county_sub <- census_county_sub_df %>%
  left_join(mtn_tick_town_number, by = c("Place" = "Location"))

# view non-matches >> census county subdivision df with mtn number df
missing <- anti_join(census_county_sub_df, mtn_tick_town_number, by = c("Place" = "Location")) %>%
                    select(Place, County, estimate)

# create labels
mtn_number_labels <- sprintf(
  "<strong>%s (2016-2020)</strong> <br/>Lyme Cases: %s",
   mtn_number_census_county_sub$Place,  mtn_number_census_county_sub$Lyme_Label
) %>% lapply(htmltools::HTML)

# define bins
bins3 <- seq(0, 160, by = 20)

# select color palettes
pal3 <- colorBin("OrRd", domain = mtn_number_census_county_sub$Lyme, bins = bins3)

# generate leaflet
leaflet(mtn_number_census_county_sub) %>%
  addProviderTiles(providers$OpenStreetMap) %>%
  addPolygons(fillColor = ~pal3(mtn_number_census_county_sub$Lyme),
              color = "black",
              weight = 1,
              fillOpacity = 0.9,
              label = ~mtn_number_labels,
              group = "Lyme") %>%
  addLegend(values = ~mtn_number_census_county_sub$Lyme, 
            group = "Lyme", 
            position = "topright", 
            title = "Lyme Disease Cases",
            opacity = 1,
            pal = pal3,
            na.label = "NA or Not Releasable") %>%
  setView(lng = -69.5, 
          lat = 45.3, 
          zoom = 6.5) 
  
```

